# Multidimensional Exploration

In this second section, we simultaneously deal with several dimensions.
Multidimensional plotting of variables is then possible.


## Selecting two dimensions with `select.dim`

When selecting two dimensions (or more) with function `select.dim`, one ends up with a multidimensional representation of the corresponding variables.
For example, in the code below, one selects dimensions `week` and `country` (in other words, dimension `media` is aggregated).
Each observation hence gives the number of `articles` published during a given `week` and talking about a given `country`.

```{r}
geomedia %>%
    select.dim (week, country) %>%
    arrange.elm (country, name) %>%
    arrange.elm (week, name) %>%
    as.data.frame ()
```

## Plotting bidimensional variables with argument `sep.dim.names` of `plot.var`

In order to plot variable `articles` according to the two selected dimensions, one can use argument `sep.dim.names` of function `plot.var` to indicate which one of the two dimensions should be used to separate the data.

```{r}
geomedia %>%
    select.dim (week, country) %>%
    arrange.elm (week, name) %>%
    filter.elm (country, name %in% c ("USA", "FRA", "DEU")) %>%
    plot.var (articles, sep.dim.names = country, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

In this context, using a point-plot (instead of a line-plot) can be useful to separate a dimension with many elements.
(Note that, below, we also use `filter.elm` on `week` to only plot data for year 2014.)

```{r}
geomedia %>%
    select.dim (week, country) %>%
    filter.elm (week, format (name, "%Y") == "2014") %>%
    arrange.elm (week, name) %>%
    top_n.elm (country, articles, 20) %>%
    arrange.elm (country, desc (articles)) %>%
    plot.var (articles, sep.dim.names = country, type = "point") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Plotting multidimensional variables


It is also possible to separate the data according to several dimensions (with not so many elements) at the same time in order to plot multidimensional variables.
For example, the plot below gives the temporal evolution of `articles` for 8 countries and 2 newspapers.

```{r}
geomedia %>%
    filter.elm (week, format (name, "%Y") == "2014") %>%
    arrange.elm (week, name) %>%
    top_n.elm (country, articles, 8) %>%
    arrange.elm (country, desc (articles)) %>%
    filter.elm (media, name %in% c ("fr_FRA_lmonde_int", "en_GBR_guardi_int")) %>%
    plot.var (articles, sep.dim.names = list (country, media), type = "point") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


