---
title: "A Tutorial for the `data.cube` Library"
author: "Robin Lamarche-Perrin"
date: "June 2019"
output:
  html_document:
    theme: united
    highlight: tango
---
	
`data.cube` is an R package for the exploration of multidimensional datasets and for the detection of statistical outliers within. It is mainly a tool for data exploration, allowing to have a first glance at it and to formulate research hypotheses to be later tested.

The package defines a new data structure called `data.cube` that can be fed with a classical `data.frame` encoding a list of numeric observations described according to several categorical dimensions. For example, in the case of Twitter data, it can be the number of tweets (numeric observation) that have been published by a given user (first dimension) about a given topic (second dimension) at a given date (third dimension). The input `data.frame` hence takes the form of a list of quadruplets (user, topic, date, number of tweets).

Statistical outliers can then be identified among the observations by first selecting some dimensions of interest, that is by subsetting or by aggregating the input dimensions. If needed, observations can also be normalised according to the marginal values along the selected dimensions, thus comparing the observed value to an expected value obtained by the uniform redistribution of the selected marginal values. Different statistical tests can then be chosen to measure the deviation between the observed and the expected values. The package finally allows to retrieve a list of positive outliers, that is observations that are significantly higher than expected.


# Table of Contents

```{r setup, include=FALSE} 
knitr::opts_chunk$set (echo = TRUE)
source ("render_toc.R")
```

```{r toc, echo=FALSE} 
render_toc ("data.cube.tuto.Rmd", toc_depth = 3)
```


# Installation

The library is available on CRAN:

```{r eval=FALSE}
install.packages ('data.cube')
```

Its source code is available on GitHub: <https://github.com/Lamarche-Perrin/data.cube>

# Authors and License

This library has been developed by researchers of the [Complex Networks](http://www.complexnetworks.fr/) team, within the [Laboratoire d'informatique de Paris 6](https://www.lip6.fr/), for the [ODYCCEUS](https://www.odycceus.eu/) project, founded by the [European Commission FETPROACT 2016-2017 program](https://ec.europa.eu/research/participants/portal/desktop/en/opportunities/h2020/calls/h2020-fetproact-2016-2017.html) under grant 732942.

Copyright © 2017-2019 Robin Lamarche-Perrin (<Robin.Lamarche-Perrin@lip6.fr>)

`data.cube` is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version. It is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GN  General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see <http://www.gnu.org/licenses/>.



# Getting Started

This tutorial assumes that you are familiar with the most known R packages of the [`tidyverse`](https://www.tidyverse.org/), in particular `tibble` operations provided by [`dplyr`](https://dplyr.tidyverse.org/) and the forward-pipe operator `%>%` provided by [`magrittr`](https://magrittr.tidyverse.org/).


## Downloading the dataset

This tutorial presents a use case of the `data.cube` library on a quite simple dataset ([downloadable here](link)) referencing the citations of country names in press articles.
This dataset has been extracted from the corpus of articles gathered by the [ANR GEOMEDIA Project](https://geomedia.hypotheses.org/).

First, import the main file of this dataset as a `data.frame` (actually, as a `tibble`):

```{r}
library (readr)
df <- read_csv ('data/articles.csv')
head (df)
```

* First column `id_media` contains standardised identifiers of selected newspapers.
* Second column `week` contains publication dates at the week level (date of the first day of the week at the `YYYY-MM-DD` format).
* Third column `id_country` contains standardised identifiers of cited countries ([ISO 3166-1 alpha-3](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3)).
* Last column `article_nb` gives the corresponding number of articles, that is the number of articles published by `id_media` during `week` and citing `id_country`.

For example, the third line of the dataset above indicates that the Venezuelan newspaper *El Universal* (`es_VEN_univer_int`) as published `9` articles talking about Libya (`LBY`) during the week starting on the 19th of May, 2014 (`2014-05-19`).
Note that the indicated number of articles is not necessarily an integer value (see for example the first line) as an article simultaneously citing `n` countries is weighted by `1/n` and then distributed among `n` lines.


## Loading the library

First, load the library:
```{r eval=FALSE}
library (data.cube)
```

```{r include=FALSE}
source ('../../src/data.cube.R')
```

## Building the cube with `as.data.cube`

Function `as.data.cube` transforms a classical `data.frame` (or `tibble`) object into a `data.cube`, that is the data structure that will then be used by the library.
One should specify which columns correspond to the cube's dimensions (in our case, the first three) and which columns correspond to the observed variables (in our case, the last one).
Note that one might also rename these dimensions and variables when transforming the `data.frame` into a `data.cube`.


```{r}
geomedia <-
    df %>%
    as.data.cube (
        dim.names = list (media = id_media, week, country = id_country),
        var.names = list (articles = article_nb)
    )
```


## What's in the cube with `summary`
Function `summary` then prints a short summary of the data contained in the resulting structure.

```{r}
geomedia %>% summary ()
```


## Back to frame with `as.data.frame`

Function `as.data.frame` transforms a `data.cube` object back into a `data.frame` object (actually, a `tibble`).

```{r}
geomedia %>% as.data.frame ()
```



# Unidimensional Exploration

In this first section, we focus on the exploration of one dimension of the `data.cube`.
Multidimensional exploration is demonstrated later.


## Selecting one dimension with `select.dim`

One first needs to select the dimension of interest using function `select.dim`.
In this example, we focus on the dataset's temporal dimension, a.k.a. dimension `week`.


```{r}
geomedia %>%
    select.dim (week) %>%
    as.data.frame ()
```

The resulting `data.cube` consists in a unidimensional data structure
where variable `articles` has been aggregated (summed) along dimensions `media` and `country`.
The resulting `data.frame` hence gives the total number of published articles corresponding to each element of dimension `week`.


## Arranging elements with `arrange.elm`

Note that above, observations have no particular order.
Function `arrange.elm` reorders elements of a given dimension according to one (or several) of their variables.
For example, the lexicographic order of their `name` (standard variable created for each dimension when instantiating the `data.cube`) which happens to also be the chronological order.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    as.data.frame ()
```

## Plotting variables with `plot.var`

Function `plot.var` then plots a variable.
Note that it returns a `ggplot` object that can hence be modified using classical tools of the visualisation library.
For example, one can use function `theme` to vertically display x-axis labels.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    plot.var (articles) +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

Several plot types are available: `bar` (above), `line` (below), and `point`.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    plot.var (articles, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Filtering elements with `filter.elm` and `top_n.elm`

Note that some observations in the plot above are surprisingly low.
(They actually correspond to technical incidents during data collection.)

Function `top_n.elm` only keeps the elements of a dimension that have the highest (or the lowest) value according to a variable.
We here plot the 10 weeks in the data that have the lowest number of published article (note the `-` in argument `n`).

```{r}
geomedia %>%
    select.dim (week) %>%
    top_n.elm (week, articles, n = -10) %>%
    arrange.elm (week, articles) %>%
    plot.var (articles) +
    theme (axis.text.x = element_text (size = 6))
```

Function `filter.elm` only keeps the elements of a dimension that fit with some criteria expressed on variables.
We can for example use it to remove such anomalous observations.

```{r}
geomedia %>%
    select.dim (week) %>%
    filter.elm (week, articles >= 2500) %>%
    arrange.elm (week, name) %>%
    plot.var (articles, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Other examples of use

Here are other examples of use of these simple operations, illustrated on the spatial dimension `country`.

One can plot the number of articles associated with the top 20 countries (arranged in a decreasing order).

```{r}
geomedia %>%
    select.dim (country) %>%
    top_n.elm (country, articles, 20) %>%
    arrange.elm (country, desc (articles)) %>%
    plot.var (articles)
```

One can filter and arrange countries according to a given subset.

```{r}
G8 <- c ("USA", "JPN", "DEU", "FRA", "RUS", "GBR", "ITA", "CAN")
geomedia %>%
    select.dim (country) %>%
    filter.elm (country, name %in% G8) %>%
    arrange.elm (country, match (name, G8)) %>%
    plot.var (articles)
```



# Multidimensional Exploration

In this second section, we simultaneously deal with several dimensions.
Multidimensional plotting of variables is then possible.


## Selecting two dimensions with `select.dim`

When selecting two dimensions (or more) with function `select.dim`, one ends up with a multidimensional representation of the corresponding variables.
For example, in the code below, one selects dimensions `week` and `country` (in other words, dimension `media` is aggregated).
Each observation hence gives the number of `articles` published during a given `week` and talking about a given `country`.

```{r}
geomedia %>%
    select.dim (week, country) %>%
    arrange.elm (country, name) %>%
    arrange.elm (week, name) %>%
    as.data.frame ()
```

## Plotting bidimensional variables with argument `sep.dim.names` of `plot.var`

In order to plot variable `articles` according to the two selected dimensions, one can use argument `sep.dim.names` of function `plot.var` to indicate which one of the two dimensions should be used to separate the data.

```{r}
geomedia %>%
    select.dim (week, country) %>%
    arrange.elm (week, name) %>%
    filter.elm (country, name %in% c ("USA", "FRA", "DEU")) %>%
    plot.var (articles, sep.dim.names = country, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

In this context, using a point-plot (instead of a line-plot) can be useful to separate a dimension with many elements.
(Note that, below, we also use `filter.elm` on `week` to only plot data for year 2014.)

```{r}
geomedia %>%
    select.dim (week, country) %>%
    filter.elm (week, format (name, "%Y") == "2014") %>%
    arrange.elm (week, name) %>%
    top_n.elm (country, articles, 20) %>%
    arrange.elm (country, desc (articles)) %>%
    plot.var (articles, sep.dim.names = country, type = "point") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Plotting multidimensional variables


It is also possible to separate the data according to several dimensions (with not so many elements) at the same time in order to plot multidimensional variables.
For example, the plot below gives the temporal evolution of `articles` for 8 countries and 2 newspapers.

```{r}
geomedia %>%
    filter.elm (week, format (name, "%Y") == "2014") %>%
    arrange.elm (week, name) %>%
    top_n.elm (country, articles, 8) %>%
    arrange.elm (country, desc (articles)) %>%
    filter.elm (media, name %in% c ("fr_FRA_lmonde_int", "en_GBR_guardi_int")) %>%
    plot.var (articles, sep.dim.names = list (country, media), type = "point") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```



# Outlier Analysis


One of the tools that the `data.cube` library offers deals with the detection of statistical outliers in multidimensional data.
In this context, variables are interpreted as multidimensional [contingency tables](https://en.wikipedia.org/wiki/Contingency_table) from which multiple marginal counts can be extracted (depending on the selected dimensions).
These marginals then allow for the design of multiple models of the observed variables, from which significant deviations are then interpreted as outliers.
In this part of the tutorial, we show how functions `compute.var.model` and `plot.outliers` allow for the computation of such models and the visualisation of the resulting deviations.

For more details regarding the formal ground of these functions, please refer to:
*Audrey Wilmet and Robin Lamarche-Perrin. (2019). Multidimensional Outlier Detection in Temporal Interaction Networks.* [arXiv:1906.02541](https://arxiv.org/abs/1906.02541)


## First example: Bidimensional outliers

This first example focuses on the evolution over year 2014 of the number of articles citing the 5 following countries: The United States, Russia, Spain, Italy, and Japan.

```{r}
geomedia_2014 <-
	geomedia %>%
	select.dim (week, country) %>%
    filter.elm (week, format (name, "%Y") == "2014") %>%
    arrange.elm (week, name)

selected_countries <- c ("USA", "RUS", "ESP", "ITA", "JPN")

geomedia_2014 %>%
    filter.elm (country, name %in% selected_countries) %>%
    plot.var (articles, sep.dim.names = country, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

### Taking into account the expected number of articles for each country

By plotting the distribution of all these observations, one could define some king of "statistically expected" range of values and then identify outliers as unexpectedly high (or unexpectedly low) values.
In the plot below, one could for example grossly characterise the distribution as a Gaussian curve (values on the left) plus some positive outliers (values on the right).

```{r}
geomedia_2014 %>%
    filter.elm (country, name %in% selected_countries) %>%
    as.data.frame () %>%
    ggplot (aes (x = articles)) +    
    geom_histogram (aes (y = ..density..), binwidth = 30, colour = "black", fill = "white") +
    geom_density (alpha = 0.2, fill = "black")
```

Yet, it seems that several Gaussian distributions are actually superimposed.
One could hence recognise that each of the five selected country has its own expected range of values, depending on its usual "media coverage" (expected number of articles per week).
In this sense, one could hence search for outliers within each of these five distributions independently.

```{r}
geomedia_2014 %>%
    filter.elm (country, name %in% selected_countries) %>%
    as.data.frame () %>%
    ggplot (aes (x = articles, color = country, fill = country)) +
    geom_histogram (aes (y = ..density..), binwidth = 30, alpha = 0.2, position = "dodge") +
    geom_density (alpha = 0.2)
```

To take into account these different ranges of values, another solution consists in normalising the observations by taking into account the expected "media coverage" of each country.
Function `compute.var.model` aims at computing such normalisation models.

This function take in argument a formula specifying the normalisation scheme that needs to be applied.
In the example below, formula `articles (week * country) ~ articles (country)` indicates that we want to model variable `articles` associated to dimensions `week` and `country` (left-hand side) by taking into account its marginal total along dimension `country` (right-hand side).
In other words, with this first model, the number of articles citing a given country during a given week is expected to be similar to *the average number of articles citing that country each week* (total number in the corpus divided by number of weeks).


```{r}
geomedia_2014 %>%
    filter.elm (country, name %in% selected_countries) %>%
    compute.var.model (articles (week * country) ~ articles (country)) %>%
    summary ()
```

Applying the input formula, function `compute.var.model` computed two new variables:

* Variable `articles.model` is the aforementioned average number of articles for each country;

* Variable `articles.deviation` provides a measure of deviation between the observed variable `article` and the model.
By default, it simply is the ratio between the two variables.
For example, `articles.deviation == 3` indicates that a given country has been three times more cited during a given week than usually.
`articles.deviation == 0.5` indicates that it has been two times less cited.

These computed variables can now be visualised.
Let's first have a look at the model by plotting `articles.model`:
It simply consists in the total number of articles, averaged among weeks, for each of the five countries.


```{r}
geomedia_2014 %>%
    compute.var.model (articles (week * country) ~ articles (country)) %>%
    filter.elm (country, name %in% selected_countries) %>%
    plot.var (articles.model, sep.dim.names = country, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

We clearly see in the plot above that each of the five countries has a specific expected "media coverage" to which observations can be compared.
To do so, we now plot variable `articles.deviation` that indicates the ratio between observed and expected values.


```{r}
geomedia_2014 %>%
    compute.var.model (articles (week * country) ~ articles (country)) %>%
    filter.elm (country, name %in% selected_countries) %>%
    plot.var (articles.deviation, sep.dim.names = country, type = "bar") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

This plot finally allows for the identification of outliers with respect to each country, relatively to its usual "media coverage".
Values are centered on 1, above are positive outliers, below are negative outliers.



### Also taking into account the expected number of articles for each week

```{r}
geomedia_2014 %>%
    select.dim (week) %>%
    plot.var (articles, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```
	
```{r}
geomedia_2014 %>%
    compute.var.model (articles (week * country) ~ articles (week) * articles (country)) %>%
    filter.elm (country, name %in% selected_countries) %>%
    plot.var (articles.deviation, sep.dim.names = country, type = "bar") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

Note that function `filter.elm` is applied after function `compute.var.model`.

<!-- ## 5.1. Data preparation -->

<!-- geomedia3 <- -->
<!--   geomedia %>% -->
<!--   select.dim (week, country) %>% -->
<!--   select.elm (country, c("USA","RUS","FRA","ITA","JPN")) %>% -->
<!--   arrange.elm (week, country) -->


<!-- ## 5.2. Compute a simple model (taking into account the global popularity of country) -->

<!-- # Raw observations -->
<!-- geomedia3 %>% plot.obs (sep.dim = country, type = "line") -->

<!-- # Raw model -->
<!-- geomedia3 %>% -->
<!--   compute.model () %>% -->
<!--   plot.obs (model, sep.dim = country, type = "line") -->

<!-- # Model taking into account 'country' marginals -->
<!-- geomedia3 %>% -->
<!--   compute.model (country) %>% -->
<!--   plot.obs (model, sep.dim = country, type = "line") -->

<!-- # Ratio between observed values and expected values -->
<!-- geomedia3 %>% -->
<!--   compute.model (country) %>% -->
<!--   plot.obs (ratio, sep.dim = country) -->


<!-- ## 5.3. Compute a more complete model (also taking into account the global activity through time) -->

<!-- geomedia3 %>% select.dim (week) %>% plot.var (articles) -->

<!-- geomedia3 %>% -->
<!--   compute.model (country, week) %>% -->
<!--   plot.obs (ratio, sep.dim = country) -->


<!-- ## 5.4. Compute significativity of  -->

<!-- geomedia3 %>% -->
<!--   compute.model (country, week, deviation.type = 'poisson') %>% -->
<!--   plot.obs (deviation, sep.dim = country) -->

<!-- geomedia3 %>% -->
<!--   compute.model (country, week, deviation.type = 'poisson') %>% -->
<!--   plot.outlier () -->

<!-- geomedia %>% -->
<!--   select.dim (country, week) %>% -->
<!--   compute.model (country, week, deviation.type = 'poisson') %>% -->
<!--   plot.outlier () -->

<!-- geomedia %>% -->
<!--   select.dim (country, week) %>% -->
<!--   compute.model (country, week, deviation.type = 'poisson') %>% -->
<!--   filter.obs (articles >= 100) %>% -->
<!--   filter.obs (ratio >= 1) %>% -->
<!--   plot.outlier () -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, country) %>% -->
<!--   compute.model (country, newspapers, deviation.type = 'poisson') %>% -->
<!--   filter.obs (articles >= 100) %>% -->
<!--   filter.obs (ratio >= 1) %>% -->
<!--   plot.outlier () -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, week) %>% -->
<!--   compute.model (week, newspapers, deviation.type = 'poisson') %>% -->
<!--   filter.obs (articles >= 100) %>% -->
<!--   filter.obs (ratio >= 1) %>% -->
<!--   plot.outlier () -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, country) %>% -->
<!--   compute.model (country, newspapers, deviation.type = 'poisson') %>% -->
<!--   filter.obs (outlier == 1) %>% -->
<!--   arrange.obs (deviation) %>% -->
<!--   as.data.frame () -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, country) %>% -->
<!--   compute.model (country, newspapers, deviation.type = 'poisson') %>% -->
<!--   filter.obs (outlier == 1) %>% -->
<!--   arrange.obs (deviation) %>% -->
<!--   as.data.frame () -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, country) %>% -->
<!--   select.elm (country, top.nb = 10) %>% -->
<!--   select.elm (newspapers, top.nb = 5) %>% -->
<!--   compute.model (country, newspapers, deviation.type = 'poisson') %>% -->
<!--   biplot.obs (newspapers, country, deviation) -->

<!-- elm.names (geomedia, week) -->

<!-- geomedia %>% -->
<!--   select.dim (newspapers, country) %>% -->
<!--   select.elm (country, top.nb = 50) %>% -->
<!--   select.elm (newspapers, top.nb = 10) %>% -->
<!--   compute.model (country, newspapers, deviation.type = 'poisson') %>% -->
<!--   biplot.obs (newspapers, country) -->



<!-- ## Temporal outliers -->

<!-- geomedia3 <- -->
<!--   geomedia %>% -->
<!--   select.elm (week, week_2014, suppress = TRUE) %>% -->
<!--   select.elm (country, G8) -->

