# An ODYCCEUS Tutorial for the `data.cube` Library {-}

`data.cube` is an R package for the exploration of multidimensional datasets and for the detection of statistical outliers within. It is mainly a tool for data exploration, allowing to have a first glance at it and to formulate research hypotheses to be later tested.

The package defines a new data structure called `data.cube` that can be fed with a classical `data.frame` encoding a list of numeric observations described according to several categorical dimensions. For example, in the case of Twitter data, it can be the number of tweets (numeric observation) that have been published by a given user (first dimension) about a given topic (second dimension) at a given date (third dimension). The input `data.frame` hence takes the form of a list of quadruplets (user, topic, date, number of tweets).

Statistical outliers can then be identified among the observations by first selecting some dimensions of interest, that is by subsetting or by aggregating the input dimensions. If needed, observations can also be normalised according to the marginal values along the selected dimensions, thus comparing the observed value to an expected value obtained by the uniform redistribution of the selected marginal values. Different statistical tests can then be chosen to measure the deviation between the observed and the expected values. The package finally allows to retrieve a list of positive outliers, that is observations that are significantly higher than expected.


## Table of Contents

```{r setup, include=FALSE}
knitr::opts_chunk$set (echo = TRUE, cache = TRUE, cache.lazy = FALSE, out.width = "\\textwidth", fig.align = "center")
source ('_render_toc.R')
```

```{r toc, echo=FALSE, cache = FALSE}
render_toc ('tutorial.Rmd', toc_depth = 3)
```


## Installation {-}

The library is not yet available on CRAN.
Its source code is available on GitHub: <https://github.com/Lamarche-Perrin/data.cube/blob/master/src/data.cube.R>

<!-- ```{r eval=FALSE} -->
<!-- install.packages ('data.cube') -->
<!-- ``` -->


## Authors and License {-}

This library has been developed by researchers of the [Complex Networks](http://www.complexnetworks.fr/) team, within the [Laboratoire d'informatique de Paris 6](https://www.lip6.fr/), for the [ODYCCEUS](https://www.odycceus.eu/) project, founded by the [European Commission FETPROACT 2016-2017 program](https://ec.europa.eu/research/participants/portal/desktop/en/opportunities/h2020/calls/h2020-fetproact-2016-2017.html) under grant 732942.

Copyright © 2017-2019 Robin Lamarche-Perrin (<Robin.Lamarche-Perrin@lip6.fr>)

`data.cube` is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version. It is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GN  General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see <http://www.gnu.org/licenses/>.



# Getting Started

This tutorial assumes that you are familiar with the most known R packages of the [`tidyverse`](https://www.tidyverse.org/), in particular `tibble` operations provided by [`dplyr`](https://dplyr.tidyverse.org/) and the forward-pipe operator `%>%` provided by [`magrittr`](https://magrittr.tidyverse.org/).


## Importing the dataset

This tutorial presents a use case of the `data.cube` library on a dataset of 30M tweets published between January 2015 and July 2016, and relating to the European migrant crisis of 2015 (request: "(e|im)?migrant(s)?").
These tweets have been extracted with [DMI-TCAT](https://github.com/digitalmethodsinitiative/dmi-tcat), a software designed by the [Digital Methods Initiative](https://wiki.digitalmethods.net/Dmi/SummerSchool2019), in the context of the [H2020 ODYCCEUS Project](https://www.odycceus.eu/).

First, import the main file of this dataset as a `data.frame` (actually, as a `tibble`):

```{r import_df, message=FALSE}
library (readr)
df.tweets <- read_csv ('data/ODYCCEUS-migrant-crisis-tweets.csv')
df.tweets
```
<!-- TODO: sample tibble -->

* First column `day` contains publication dates of the tweets (`YYYY-MM-DD` format).

* Second column `user` contains account names of users that have published the tweets.
Accounts with less than 1000 followers have been anonymised (for example `user12345`).

* Third column `type` indicates the type of tweet: `tweet`, `reply`, or `retweet`.

* Fourth column `to_user` contains the account names of users that have been replied to (if `type == 'reply'`), or that have been retweeted (if `type == 'retweet'`), or `NULL` in the case of an original tweet (if `type == 'tweet'`).

* Fifth column `hashtag` indicates eventual hashtags within the tweets (`NULL` if no hashtag).

* Sixth column `count` indicates the corresponding number of tweets, that is the number of tweets published by `user` during `day` of `type` relating `to_user` and containing `hashtag`.

* Seventh column `weight` takes into account the fact that several hashtags might appear in the same tweet, and so that such a tweet might hence be counted several times.
In this regard, a tweet containing `n` different hashtags is weighted by `1/n` and then distributed among `n` lines.
Summing this last columns hence gives the exact number of tweets (contrary to the `count` column).

<!-- For example, the third line of the dataset above indicates that on the 1st of January 2015 (`2015-01-01`) user `055Douglas` retweeted (`retweet`) a tweet from user `mcspocky` containing hashtag `Immigration`, and that it happened only once (`count == 1`). -->


## Loading the library

First, load the library:
```{r source_lib1, eval=FALSE}
source ('data.cube.R')
# or library (data.cube) when it will be available as a proper R package
```

```{r source_lib2, include=FALSE, eval=TRUE}
source ('../../src/data.cube.R')
```

## Building the cube with `as.data.cube`

Function `as.data.cube` transforms a classical `data.frame` (or `tibble`) object into a `data.cube`, that is the data structure that will then be used by the library.
One should specify which columns correspond to the cube's dimensions (in our case, the first five) and which columns correspond to the observed variables (in our case, the last one).
Note that one might also rename these dimensions and variables when transforming the `data.frame` into a `data.cube`.


```{r as_dc}
dc.tweets <-
    df.tweets %>% as.data.cube (
               dim.names = list (day, from_user, type, to_user, hashtag),
               var.names = list (tweets = weight, count),
               sup.dim.names = list (user = list (from_user, to_user))
           )
```


## What's in the cube with `summary`
Function `summary` then prints a short summary of the data contained in the resulting structure.

```{r summary_dc}
dc.tweets %>% summary (verbose = TRUE)
```


## Back to frame with `as.data.frame`

Function `as.data.frame` transforms a `data.cube` object back into a `data.frame` object (actually, a `tibble`).

```{r as_df}
dc.tweets %>% as.data.frame ()
```


# When do we talk about what?

This first `data.cube` and its 5 dimensions are a bit tedious for a first analysis.
The library provide simple tools for subsetting the data and begin the exploration along fewer dimensions.
In this section, we hence focus on the analysis of the temporal popularity of hashtags by only selecting the `day` and `hashtag` dimensions: "*When* do we talk about *what*?"


## Filtering elements, renaming variables, selecting dimensions, and arranging elements

In the script below:

* `filter.elm (type, name == 'tweet')` filters out elements of dimension `type` which are not tweets (*i.e.*, filters out replies and retweets);
* `rename.var (tweets = documents)` renames variable `documents` with new name `tweets`;
* `select.dim (day, hashtag)` selects dimensions `day` and `hashtag` for the following analysis (*i.e.*, aggregate the data along other dimensions `from_user`, `to_user`, and `type`);
* `arrange.elm.by.name ()` arranges elements of all dimensions according to their name (in lexicographic order).


```{r build_dc.day.hashtag}
dc.day.hashtag <-
    dc.tweets %>%
    filter.elm (type, type == 'tweet') %>%
    select.dim (day, hashtag) %>%
    arrange.elm.by.name ()

dc.day.hashtag %>% summary ()

dc.day.hashtag %>% as.data.frame ()
```
<!-- TODO: randomize -->

## First plot: Focusing on the temporal dimension

We now have a two-dimensional `data.cube` containing all hashtags through time appearing in the corpus' tweets.
First, one can select dimension `day` and plot variable `tweets` with function `plot.var` to have an aggregated representation of the number of published tweets each day (regardless of the hashtags they contain).

```{r plot_day}
dc.day.hashtag %>%
    select.dim (day) %>%
    ggplot.var (tweets)
```


Note that function `group.day.elm` easily allows to change the temporal resolution of a `data.cube` containing a dimension `day`.
A parameter indicates the name of the replacing temporal dimension: `dim.name %in% c ('week', 'month', 'year')`.

Also note that function `plot.var` returns a `ggplot` object that can hence be customised accordingly.
Below, we use `scale_x_date` to change the date labels on the x-axis and `ggtitle` to add a title to the plot.

```{r plot_week}
dc.day.hashtag %>%
    select.dim (day) %>%
    group.day.elm (week) %>%
    ggplot.var (tweets) +
    ggtitle ("Number of tweets through time")
```

## Second plot: Looking at hashtags

We now look at the second dimension of our `data.cube`, that is the dimension of hashtags (regardless of publication dates).
In the script below:

* `filter.elm (hashtag, ! name %in% hashtags.to.remove, name != 'NULL')` filters out hashtags that belongs to a user-defined list (those corresponding to keywords that were used to collect the initial corpus) and also removes hashtag `NULL` (corresponding to tweets in the corpus that do not contain any hashtag).

* `top_n.elm (hashtag, tweets, n = 20)` retains the 20 hashtags that appear in the most tweets;

* `arrange.elm (hashtag, desc (tweets))` arranges the retained hashtag with respect to the number of tweets in which they appear (in decreasing order).


```{r plot_hashtags}
hashtags.to.remove <- c ('NULL', 'emigrant', 'emigrants', 'immigrant', 'immigrants', 'migrant', 'migrants', 'Emigrant', 'Emigrants', 'Immigrant', 'Immigrants', 'Migrant', 'Migrants')

dc.hashtag <-
    dc.day.hashtag %>%
    select.dim (hashtag) %>%
    filter.elm (hashtag, ! hashtag %in% hashtags.to.remove) %>%
    top_n.elm (hashtag, tweets, n = 20) %>%
    arrange.elm (hashtag, desc (tweets))

dc.hashtag %>% ggplot.var (tweets)
```


## Outlier detection: Looking at popular hashtags through time

The `data.cube` library offers simple tools for outlier detection in multidimensional data.
Let's look for dates when given hashtags are particularly popular (*cf.* country names in script below).

We first build a subset of the previous `data.cube` by selecting 7 hashtags of interest.

```{r outlier_detection_1}
hashtags.of.interest <- c ('Iran', 'Iraq', 'Kosovo', 'Eritrea', 'Nigeria', 'Lybia')

dc.day.hashtag.subset <-
    dc.day.hashtag %>%
    filter.elm (hashtag, hashtag %in% hashtags.of.interest) %>%
    arrange.elm (hashtag, desc (tweets)) %>%
    group.day.elm (week)
```

We now use function `plot.var` with parameter `sep.dim.names = hashtag` to plot 5 lines, each corresponding to the number of tweets in which the 5 selected hashtags appear.

```{r outlier_detection_2}
dc.day.hashtag.subset %>%
    ggplot.var (tweets, sep.dim.names = hashtag, type = 'line')
```

Finding outliers in the plot above is problematic.
Because hashtag `Iran` is much more present than other ones, most curves look quite low.
Indeed, if one get the aggregated tweet number of selected hashtags (regardless of publication dates), one notice that they globally have quite different levels of popularity on the whole corpus (*e.g,* `Iran` appears in 12 times more tweets than `Kosovo`).

```{r outlier_detection_3}
dc.day.hashtag.subset %>%
    select.dim (hashtag) %>%
    as.data.frame
```

Finding outliers hence requires to first define an "expectation model" to which observations must then be compared.
There are multiple ways of defining such a model.
For a start, one can take into account the aggregated values above to build expected values.

In the script below, function `mutate.var.model` applied on variable `tweets` with parameter `tweets (hashtags)` defines a model which (i)&nbsp;projects the variable onto dimension `hashtag` (*i.e.*, aggregates dimension `week`), (ii)&nbsp;get the resulting tweet number for each hashtag, (iii)&nbsp;uniformly distributes this aggregated variable within dimension `week` to obtain an average value.
It results in the computation of a new variable, `tweets.model <- tweets (hashtag) / elm.nb (week)`, which is displayed in the following plot.


```{r outlier_detection_4, message=FALSE}
dc.day.hashtag.subset %>%
    complete.elm () %>%
    mutate.var.model (tweets, tweets (hashtag)) %>%
    ggplot.var (tweets.model, sep.dim.names = hashtag, type = 'line')
```

Using this computed model, function `mutate.var.deviation` then computes a third variable, `tweets.deviation`, which indicates to which extent observations deviate from the model.
With no additional parameter, the deviation is given by a simple ratio `tweets.deviation <- tweets / tweets.model`.

```{r outlier_detection_5, message=FALSE}
dc.day.hashtag.subset %>%
    mutate.var.model (tweets, tweets (hashtag)) %>%
    mutate.var.deviation (tweets, deviation.type = "ratio") %>%
    ggplot.var (tweets.deviation, sep.dim.names = hashtag, type = 'bar')
```

One can now see in the plot above that hashtags that all hashtags show outlier values, even those that are globally less popular than others (such as `Kosovo` in a week of May 2016, which has been cited 12 times more during that week than the averaged value).

Note that ratio does not account for the significativity of a deviation.
Indeed, observing a value of `4` when expecting `2` is equivalent to observing a value of `400` when expecting `200`, although the former might easily be caused by insignificant statistical variations while the latter is much more reliable.
Function `mutate.var.deviation` can hence be adjusted with parameter `deviation.type %in% c ('ratio', 'poisson', 'chisq', 'KLdiv')` to use other deviation measures which take into account the significativity of such deviations.

For example, in the script below, `deviation.type = 'chisq'` indicates to use a measure based on the Chi-squared test: `tweets.deviation <- (tweets - tweets.model)^2 / tweets.model`.

```{r outlier_detection_6, message=FALSE}
dc.day.hashtag.subset %>%
    mutate.var.model (tweets, tweets (hashtag)) %>%
    mutate.var.deviation (tweets, deviation.type = 'chisq') %>%
    ggplot.var (tweets.deviation, sep.dim.names = hashtag, type = 'bar')
```

It then appears in the plot above that `Kosovo` indeed shows less outliers that computed before when accounting for the statistical significativity of such outliers.


<!-- ```{r outlier_detection_7, message=FALSE} -->
<!-- dc.day.hashtag.subset %>% -->
<!--     mutate.var.model (tweets (week * hashtag), tweets (hashtag)) %>% -->
<!--     mutate.var.deviation (tweets (week * hashtag), deviation.type = 'chisq') %>% -->
<!--     arrange.elm (list (week, hashtag), desc (tweets.deviation)) %>% -->
<!--     as.data.frame -->
<!-- ``` -->
	

# Who is retweeting whom?

This next section deals with another subset of the `data.cube` dimensions, that is the retweet relation between users: "*Who* is retweeting *whom*?"


## Joining cubes to exploit other variables

Because out initial `data.cube` contains `dc.tweets %>% elm.nb (user) = 4738410` users, which is quite huge for a first exploration, we choose to reduce this number by only retaining most influential users.
To do so, we exploit another variable, that is the number of followers of each user.

This variable is first imported from another `data.frame`.

```{r import_df_user, message=FALSE}
df.users <- read_csv ('data/ODYCCEUS-migrant-crisis-users.csv')
df.users
```

One can now build a uni-dimensional `data.cube` from the imported `data.frame` and join it with our first `data.cube` with function `join`.
It results in a `data.cube` of 5 dimensions (as before) with a total of 4 variables (2 of them being defined on dimension `user` only).

```{r add_user_info}
dc.users <-
    df.users %>%
    as.data.cube (
        dim.names = user,
        var.names = list (tweetcount, followercount)
    )

dc.tweets <- dc.tweets %>% join (dc.users)
dc.tweets %>% summary ()
```

We now check the number of users that have more than 5000 followers.

```{r subset_for_exp2}
dc.tweets %>% filter.elm (user, followercount >= 5000) %>% elm.nb (user)
```

We finally use function `group.elm` to aggregate together all users that have less than 5000 followers within a new element named `OTHER_USERS`.
This drastically reduces the size of dimension `user` while keeping marginal information about the variables (contrary to function `filter.elm` which would have simply removed these users, along with all associated observations).

```{r build_exp2}
dc.from_user.to_user <-
    dc.tweets %>%
    filter.elm (type, type == 'retweet') %>%
    rename.var (retweets = tweets) %>%
    group.elm (user, OTHER_USERS, followercount < 5000) %>%
    select.dim (from_user, to_user)    

dc.from_user.to_user %>% summary ()
```


## Two-dimensional plotting

When projected onto dimension `user`, variable `retweets` now leads to two different variables: `retweets.from_user`, which corresponds to the number of retweets published by user, and `retweets.to_user`, which corresponds to the retweet number of tweets published by a given user.
The former could be interpreted as the *activity* of users (as broadcasters) and the latter as the *popularity* of users (as authors).

In the script below, function `biplot.var` allows to plot these two variables one against the other to explore the activity and popularity levels of a subset of users.

```{r plot_users}
dc.from_user.to_user %>%
    select.dim (user) %>%
    filter.elm (user, retweets.from_user >= 500, retweets.to_user >= 500, user != 'OTHER_USERS') %>%
    ggplot.vars (retweets.from_user, retweets.to_user, log = 'xy')
```

## Exporting the cube to `igraph` for network analysis

Function `as.igraph` allows to transform a two-dimensional `data.cube` into an `igraph` object, which can then serve for network analysis.

```{r get_igraph_user}
igraph.user <-
    dc.from_user.to_user %>%
    filter.elm (user, user != 'OTHER_USERS') %>%
    as.igraph (from_user, to_user, edge.weight = retweets)
```

Because a graph of `vcount (igraph.user) == 165875` nodes is difficult to plot, in the script below, we use function `cluster_louvain` from the `igraph` library to group users into communities (*i.e.*, densely connected groups of nodes, see <https://igraph.org/r/doc/cluster_louvain.html> for detailed documentation).
The resulting data structure hence associates a community index to each user.

```{r compute_communities}
communities <-
    igraph.user %>%
    as.undirected (mode = 'collapse', edge.attr.comb = sum) %>%
    cluster_louvain ()

membership (communities) %>% sample (4)
```

<!-- ## small.igraph.user <- -->
<!-- ##     igraph.user %>% -->
<!-- ##     delete.vertices (communities [sizes (communities) < 100] %>% unlist %>% unname) %>% -->
<!-- ##     delete.vertices (which (degree(igraph.user) < 100)) -->

<!-- ## edge.weight <- E(small.igraph.user)$weight %>% rescale (c (1, 10)) -->

<!-- ## plot (small.igraph.user, edge.width = edge.weight, vertex.color = rainbow (length (communities), alpha=0.6) [communities$membership]) -->

In order to label these communities, we also use function `pagerank` from the `igraph` library to measure the level of importance of each user in the network (see <https://igraph.org/r/doc/page_rank.html> for detailed documentation).

```{r compute_pagerank}
pagerank <- page_rank (igraph.user)

pagerank$vector %>% sample (4)
```
<!-- pagerank$vector %>% sort (decreasing = TRUE) %>% head -->

Using function `join`, we add the two newly computed variables to our `data.cube` for later use.

```{r join_communities}
dc.from_user.to_user <-
    dc.from_user.to_user %>%
    join (
        membership (communities) %>% as.data.cube (user),
        pagerank$vector %>% as.data.cube (user, pagerank)
    )

dc.from_user.to_user %>% summary
```

These variables are now exploited to aggregate dimension `user` according to variable `community` with function `group.elm.by.var`.
We also specify two new variables that are computed during the aggregation:
* `label` is the concatenated names of the 5 most important users (according to variable `pagerank`) in each community;
* `size` simply counts the number of users in each community.

```{r group_by_community}
dc.from_community.to_community <-
    dc.from_user.to_user %>%
    group.elm.by.var (
        user, community,
        label = user [order (desc (pagerank)) [1:4]] %>% paste0 (collapse = '\n'),
        size = n()
    ) %>%
    rename.dim (from_community = from_user, to_community = to_user)

dc.from_community.to_community %>% summary
```

```{r plot_communities}
dc.from_community.to_community %>%
    select.dim (community) %>%
    filter.elm (community, size > 1000, community != 'OTHER_USERS') %>%
    mutate.var (community, community = label) %>%
    ggplot.vars (size, retweets.to_community, log = 'xy')
```

As shown in the plot below (computed with the `ggplot` library), there are many small communities and few big ones.

```{r communities_distribution, message=FALSE, warning=FALSE}
dc.from_community.to_community %>%
    as.data.frame (community) %>%
    ggplot (aes (x = size)) +
    geom_histogram () + scale_x_log10 () + scale_y_log10 ()
```

We hence extract community containing more than 1000 users (with variable `size`) and export the resulting aggregated network as an `igraph` object.

```{r get_display_community}
dc.from_community.to_community %>%
    filter.elm (community, size >= 1000) %>%
    arrange.elm (community, desc (size)) %>%
    mutate.var (community, community = label) %>%
    ggplot.var (retweets, sep.dim.names = from_community, type = "point")
```

<!-- ```{r get_display_community} -->
<!-- dc.from_community.to_community %>% -->
<!--     filter.elm (community, size >= 1000) %>% -->
<!--     arrange.elm (community, desc (size)) %>%  -->
<!--     mutate.var (community, name = label) %>% -->
<!--     filter.elm (list (from_community, to_community), from_community == to_community) %>% -->
<!--     mutate.vars.outlier (retweets (from_community * to_community), retweets (from_community) * retweets (to_community)) %>% -->
<!--     plot.var (retweets, sep.dim.names = from_community, type = "point") -->
<!-- ``` -->

```{r get_igraph_community}
igraph.community <-
    dc.from_community.to_community %>%
    filter.elm (community, size >= 1000) %>%
    mutate.var (community, community = label) %>%
    as.igraph (from_community, to_community, vertex.size = size, edge.weight = retweets)
```

Finally, one can display the result (here using the `ForceAtlas2` layout algorithm).

```{r display_communities, fig.height = 10, fig.width = 10}
library (ForceAtlas2)
set.seed (4)

vertex.size <- 100 * V(igraph.community)$size / max (V(igraph.community)$size)
edge.weight <- 20 * E(igraph.community)$weight / max (E(igraph.community)$weight)

vertex.coordinate <-
    igraph.community %>%
    layout.forceatlas2 (k = 200, gravity = 5000, iterations = 2000, plotstep = 0, linlog = TRUE)

igraph.community %>%
    plot (
        layout = vertex.coordinate,
        vertex.size = vertex.size,
        edge.curved = 0.1,
        edge.width = edge.weight,
    )
```

<!-- vertex.label.cex = 0.6, -->
<!-- edge.arrow.width = 0.5, -->
<!-- edge.arrow.size = 0.5 -->

# What is related to what?

```{r output_topicmodel}
dc.month.user <-
    dc.tweets %>%
    filter.elm (type, type %in% c ('tweet', 'reply')) %>%
    filter.elm (user, followercount >= 1000 | user == 'NULL') %>%
    filter.elm (hashtag, hashtag != 'NULL') %>%
    transmute.var (text = hashtag, size = 1) %>%
    group.day.elm (month) %>%
    select.dim (month, from_user)
    
dc.month.user %>% as.data.frame ()

dc.month.user %>%
    as.json () %>%
    write ('data/ODYCCEUS-migrant-crisis-hashtags-1000.json')
```
