---
title: "Multidimensional Outlier Explorer" 
output: 
  flexdashboard::flex_dashboard:
    css: bootstrap.css
    social: [ "menu" ]
    logo: OdycceusSymbolVerySmall.svg
    
runtime: shiny
---



```{r setup, include=FALSE}
#### OBSERVATORY INTERFACE PREMISES ####
knitr::opts_chunk$set(warning = TRUE)
require(rmarkdown)
require(markdown)
require(knitr)
require(shiny)
require(shinyjs)
require(flexdashboard)
require(shinythemes)
require(shinyWidgets)
require(shinyhelper)
require(htmltools)
require(tidyverse)
require(DT)
require(base64enc) # fix for twitter oauth in shinyapps.io
require(jsonlite)
require(httr)
require(plumber)
require(future)
future::plan(multiprocess)

#RUN exampleAPI.R
## RtweeetAPI = plumber::plumb(file= "apis/example/APIexample.R")  # Where the file describing the APIexample.R is file is located
## future::future(RtweeetAPI$run(port=5685,host="127.0.0.1", swagger=TRUE))#run exampleAPI in another R session
                # do other computationally expensive stuff here

```



Home
=====================================
<style>
body {
text-align: justify}
</style>


<center>
<img src="OdycceusIntegratedSymbolMedium.svg" width=18%>
</center>

#### Welcome 




Showcase 1 {data-navmenu="Showcases"}
================================================================

### Title Showcase 1




Iframe Example {data-navmenu="Observatory"}
===========================================================

```{r iframe, include=TRUE}
shinyApp(
ui = {fluidPage(titlePanel("Causation Tracker"),
 
                mainPanel(fluidRow(
                  htmlOutput("Babel")
                )
                )
)
},
function(input, output) {
  output$Babel<- renderUI({tags$iframe(src="http://127.0.0.1:5000", frameborder="0", scrolling="yes", width="100%", height=700)})

 }
)
```


Shiny App Example {data-navmenu="Observatory"}
===========================================================

Column {.tabset}
-----------------------------------------------------------------------

### ShinyApp


```{r app example, echo=FALSE}
shiny::shinyAppDir("apps/example",
  options = list(width = "100%", height = 800)
)
```

### Tutorial for using app example 

Dear internaut welcome to a shiny application developed as illustrative example for the penelope platform of the Odysseus project.
<br>
This Shiny application allows you to download and analyse Reddit data hrough a user friendly analysis pipeline.

<!-- insert image of the process here -->
[TODO]()

<br>
You can select the subreddit of the data to download through the query parameter (Subreddit:)
<br>
The data accessed through this Shiny APP will be downloaded from the endpoint of the  exampleAPI. 
[TODO]()
<br>


About
================================================================


#### Title
