---
title: "Multidimensional Outlier Explorer" 
output: 
  flexdashboard::flex_dashboard:
    css: bootstrap.css
    social: [ "menu" ]
    logo: pics/OdycceusSymbolVerySmall.svg
runtime: shiny
---


```{r setup, include = FALSE}
## LOAD PACKAGES
## knitr::opts_chunk$set (warning = TRUE)
## require (rmarkdown)
## require (markdown)
## require (knitr)
## require (shiny)
## require (shinyjs)
## require (flexdashboard)
## require (shinythemes)
## require (shinyWidgets)
## require (shinyhelper)
## require (htmltools)
## require (tidyverse)
## require (DT)
## require (base64enc) # fix for twitter oauth in shinyapps.io
## require (jsonlite)
## require (httr)
## require (plumber)
## require (future)
## future::plan (multiprocess)

## RUN APIS IN PARALLEL PROCESSES
## RtweeetAPI = plumber::plumb (file = "apis/example/APIexample.R")
## future::future (RtweeetAPI$run (port = 5685, host = "127.0.0.1", swagger = TRUE))
```


# Home

## First column

#### Welcome


This interface aims at the exploration of multidimensional datasets and at the detection of statistical outliers within.
It is mainly a tool for data exploration, allowing to have a first glance at it and to formulate research hypotheses to be later tested.

The package defines a new data structure called data.cube that can be fed with a classical data.frame encoding a list of numeric observations described according to several categorical dimensions. For example, in the case of Twitter data, it can be the number of tweets (numeric observation) that have been published by a given user (first dimension) about a given topic (second dimension) at a given date (third dimension). The input data.frame hence takes the form of a list of quadruplets (user, topic, date, number of tweets).

Statistical outliers can then be identified among the observations by first selecting some dimensions of interest, that is by subsetting or by aggregating the input dimensions. If needed, observations can also be normalised according to the marginal values along the selected dimensions, thus comparing the observed value to an expected value obtained by the uniform redistribution of the selected marginal values. Different statistical tests can then be chosen to measure the deviation between the observed and the expected values. The package finally allows to retrieve a list of positive outliers, that is observations that are significantly higher than expected.

For more details regarding the formal grounds of this work, please refer to:
Audrey Wilmet and Robin Lamarche-Perrin. Multidimensional Outlier Detection in Temporal Interaction Networks. Research Report, arXiv:1906.02541, June 2019. <https://arxiv.org/abs/1906.02541>


## Second column

#### How to use

Explain here how to use the explorer.


## Third column

<center><img src="pics/OdycceusIntegratedSymbolMedium.svg" width=50%></center>

#### Authors

This interface has been developed the [Complex Networks](http://www.complexnetworks.fr/) team at the [Computer Science Laboratory of Paris 6](https://www.lip6.fr/).
It is part of the [ODYCCEUS](https://www.odycceus.eu/) project founded by the [European Commission FETPROACT 2016-2017 program](https://ec.europa.eu/research/participants/portal/desktop/en/opportunities/h2020/calls/h2020-fetproact-2016-2017.html) under grant 732942.

#### Sources

This interface builds on `data.cube`, an `R` library for the easy processing a multidimensional data.
Sources are freely available on GitHub: <https://github.com/Lamarche-Perrin/data.cube>

#### License

Copyright © 2017-2019 Robin Lamarche-Perrin (<Robin.Lamarche-Perrin@lip6.fr>)

`data.cube` is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version. It is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GN  General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see <http://www.gnu.org/licenses/>.




<!-- # Iframe Example {data-navmenu="Observatories"} -->

<!-- ```{r iframe, include = TRUE} -->
<!-- shinyApp ( -->
<!--     ui = { fluidPage ( -->
<!--                titlePanel ("Causation Tracker"), -->
<!--                mainPanel (fluidRow (htmlOutput ("Babel"))) -->
<!--            ) -->
<!--     }, -->
<!--     function (input, output) { -->
<!--         output$Babel <- renderUI ({ tags$iframe (src = "http://127.0.0.1:5000", frameborder = "0", scrolling = "yes", width = "100%", height = 700) }) -->
<!--     } -->
<!-- ) -->
<!-- ``` -->


# Outlier Explorer {data-navmenu="Observatories"}

## Column {.tabset}

### ShinyApp

```{r, echo = FALSE}
shiny::shinyAppDir ("outlier-explorer", options = list (width = "100%", height = 800)
)
```


### Tutorial for using app example 

Dear internaut welcome to a shiny application developed as illustrative example for the penelope platform of the Odysseus project.
<br>
This Shiny application allows you to download and analyse Reddit data hrough a user friendly analysis pipeline.

<!-- insert image of the process here -->
[TODO]()

<br>
You can select the subreddit of the data to download through the query parameter (Subreddit:)
<br>
The data accessed through this Shiny APP will be downloaded from the endpoint of the  exampleAPI. 
[TODO]()
<br>




# Geomedia Outliers {data-navmenu="Showcases"}

### Title

Add text here.


# Guardian Outliers {data-navmenu="Showcases"}

### Title

Add text here.


# Twitter Outliers {data-navmenu="Showcases"}

### Title

Add text here.

