# Getting Started

This tutorial assumes that you are familiar with the most known R packages of the [`tidyverse`](https://www.tidyverse.org/), in particular `tibble` operations provided by [`dplyr`](https://dplyr.tidyverse.org/) and the forward-pipe operator `%>%` provided by [`magrittr`](https://magrittr.tidyverse.org/).


## Downloading the dataset

This tutorial presents a use case of the `data.cube` library on a quite simple dataset ([downloadable here](link)) referencing the citations of country names in press articles.
This dataset has been extracted from the corpus of articles gathered by the [ANR GEOMEDIA Project](https://geomedia.hypotheses.org/).

First, import the main file of this dataset as a `data.frame` (actually, as a `tibble`):

```{r}
library (readr)
df <- read_csv ('data/articles.csv')
head (df)
```

* First column `id_media` contains standardised identifiers of selected newspapers.
* Second column `week` contains publication dates at the week level (date of the first day of the week at the `YYYY-MM-DD` format).
* Third column `id_country` contains standardised identifiers of cited countries ([ISO 3166-1 alpha-3](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3)).
* Last column `article_nb` gives the corresponding number of articles, that is the number of articles published by `id_media` during `week` and citing `id_country`.

For example, the third line of the dataset above indicates that the Venezuelan newspaper *El Universal* (`es_VEN_univer_int`) as published `9` articles talking about Libya (`LBY`) during the week starting on the 19th of May, 2014 (`2014-05-19`).
Note that the indicated number of articles is not necessarily an integer value (see for example the first line) as an article simultaneously citing `n` countries is weighted by `1/n` and then distributed among `n` lines.


## Loading the library

First, load the library:
```{r eval=FALSE}
library (data.cube)
```

```{r include=FALSE}
source ('../../src/data.cube.R')
```

## Building the cube with `as.data.cube`

Function `as.data.cube` transforms a classical `data.frame` (or `tibble`) object into a `data.cube`, that is the data structure that will then be used by the library.
One should specify which columns correspond to the cube's dimensions (in our case, the first three) and which columns correspond to the observed variables (in our case, the last one).
Note that one might also rename these dimensions and variables when transforming the `data.frame` into a `data.cube`.


```{r}
geomedia <-
    df %>%
    as.data.cube (
        dim.names = list (media = id_media, week, country = id_country),
        var.names = list (articles = article_nb)
    )
```


## What's in the cube with `summary`
Function `summary` then prints a short summary of the data contained in the resulting structure.

```{r}
geomedia %>% summary ()
```


## Back to frame with `as.data.frame`

Function `as.data.frame` transforms a `data.cube` object back into a `data.frame` object (actually, a `tibble`).

```{r}
geomedia %>% as.data.frame ()
```

