# Unidimensional Exploration

In this first section, we focus on the exploration of one dimension of the `data.cube`.
Multidimensional exploration is demonstrated later.


## Selecting one dimension with `select.dim`

One first needs to select the dimension of interest using function `select.dim`.
In this example, we focus on the dataset's temporal dimension, a.k.a. dimension `week`.


```{r}
geomedia %>%
    select.dim (week) %>%
    as.data.frame ()
```

The resulting `data.cube` consists in a unidimensional data structure
where variable `articles` has been aggregated (summed) along dimensions `media` and `country`.
The resulting `data.frame` hence gives the total number of published articles corresponding to each element of dimension `week`.


## Arranging elements with `arrange.elm`

Note that above, observations have no particular order.
Function `arrange.elm` reorders elements of a given dimension according to one (or several) of their variables.
For example, the lexicographic order of their `name` (standard variable created for each dimension when instantiating the `data.cube`) which happens to also be the chronological order.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    as.data.frame ()
```

## Plotting variables with `plot.var`

Function `plot.var` then plots a variable.
Note that it returns a `ggplot` object that can hence be modified using classical tools of the visualisation library.
For example, one can use function `theme` to vertically display x-axis labels.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    plot.var (articles) +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```

Several plot types are available: `bar` (above), `line` (below), and `point`.

```{r}
geomedia %>%
    select.dim (week) %>%
    arrange.elm (week, name) %>%
    plot.var (articles, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Filtering elements with `filter.elm` and `top_n.elm`

Note that some observations in the plot above are surprisingly low.
(They actually correspond to technical incidents during data collection.)

Function `top_n.elm` only keeps the elements of a dimension that have the highest (or the lowest) value according to a variable.
We here plot the 10 weeks in the data that have the lowest number of published article (note the `-` in argument `n`).

```{r}
geomedia %>%
    select.dim (week) %>%
    top_n.elm (week, articles, n = -10) %>%
    arrange.elm (week, articles) %>%
    plot.var (articles) +
    theme (axis.text.x = element_text (size = 6))
```

Function `filter.elm` only keeps the elements of a dimension that fit with some criteria expressed on variables.
We can for example use it to remove such anomalous observations.

```{r}
geomedia %>%
    select.dim (week) %>%
    filter.elm (week, articles >= 2500) %>%
    arrange.elm (week, name) %>%
    plot.var (articles, type = "line") +
    theme (axis.text.x = element_text (angle = 90, size = 6))
```


## Other examples of use

Here are other examples of use of these simple operations, illustrated on the spatial dimension `country`.

One can plot the number of articles associated with the top 20 countries (arranged in a decreasing order).

```{r}
geomedia %>%
    select.dim (country) %>%
    top_n.elm (country, articles, 20) %>%
    arrange.elm (country, desc (articles)) %>%
    plot.var (articles)
```

One can filter and arrange countries according to a given subset.

```{r}
G8 <- c ("USA", "JPN", "DEU", "FRA", "RUS", "GBR", "ITA", "CAN")
geomedia %>%
    select.dim (country) %>%
    filter.elm (country, name %in% G8) %>%
    arrange.elm (country, match (name, G8)) %>%
    plot.var (articles)
```
